---
title: "Dynamic `issueirt`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{dynamic_issueirt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  fig.width = 5,
  fig.height = 5
)
```


## Overview

This vignette provides a tutorial on how to use the dynamic extension of `IssueIRT` model.
Please refer to the `vignette("using_issueirt")` for the basic usage of the `IssueIRT` model.
Further details on dynamic model can be found in [@issueirt].

In this vignette, we will be revisiting the validation study of [@issueirt]: The Rise of Populism and Sectional Division in the 1890s (The 52nd to 54th U.S. House of Representatives).
**Please do not make substantive conclusions based on the results here since the sample sizes are kept small for illustration purposes.**

### Installation

You can install `issueirt` package from GitHub using the `devtools` package:

```{r setup, eval=FALSE}
# install.packages("devtools")
devtools::install_github("sooahnshin/issueirt", build_vignettes = TRUE)
```

## Step 1: Prepare the data

### The 52nd to 54th U.S. House of Representatives (1891 - 1896)

**Data sources:**

- Roll call votes: Lewis, Jeffrey B., Keith Poole, Howard Rosenthal, Adam Boche, Aaron Rudkin, and Luke Sonnet. (2024) \emph{Voteview: Congressional Roll-Call Votes Database.} \url{https://voteview.com/}
- Issue labels: Bateman, David A, Ira Katznelson and John S Lapinski. (2022) "Issue codings for roll call in the U.S. House of Representatives and Senate between 1877 and 2011" \emph{American Institutions Project.} \url{http://www.davidalexbateman.net/congressional-data.html.}

We will be using the following data:

```{r southern_data}
library(issueirt)

data("us1890s_issue")
data("us1890s_legis")
data("us1890s_bills")
data("us1890s_votes")
```

Please refer to documentation for more details on the data. Also, please refer to `data-raw/us_southern.R` for the data preparation process.

Following code will prepare miscellaneous data for the visualization:

```{r southern_data_prep}
## to visualize the data
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(tibble)
library(purrr)
theme_set(theme_classic(base_size = 15) + theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5)))

southern <- c("MD", "DE", "VA", "WV", "KY", "NC", "SC", "TN", "GA", "AL", "FL", "MS", "LA", "AR", "MO", "OK", "TX")
breaks.group <- c("Southern Democrat", "Republican", "Democrat", "Ind. Democrat", "Populist", "Silver")
values.shape <- c(19, 15, 19, 19, 17, 17)
values.color <- c("#e36934", "#DE0100", "#1405BD", "#635FBD", "#62bd6d", "#777777")
years <- map(52:54, ~get_congress_years(.))
years <- paste0(str_extract(years[[1]][1], "[0-9][0-9][0-9][0-9]"), " ", str_extract(years[[length(years)]][1], "- [0-9][0-9][0-9][0-9]"))
irt.title <- paste("Ideal Points Measured with BIRT\nThe U.S. House of Representatives\n(", years, ")", sep = "")
issueirt.title <- paste("Ideal Points Measured with Issue IRT\nThe U.S. House of Representatives\n(", years, ")", sep = "")
us1890s_legis <- map(1:length(us1890s_legis), ~us1890s_legis[[.x]] |> 
                       mutate(group = ifelse(state_abbrev %in% southern & party_name == "Democrat", "Southern Democrat", party_name)))
```

### (Optional) Filter votes and legislators

In this example, we will be using non-unanimous votes and legislators who have cast at least 20 votes.
We can use `filter_votes()` function to filter votes and legislators:

```{r filter_votes}
## filter votes
n_terms <- length(us1890s_votes)
us1890s_rc <- map(1:n_terms, ~pscl::rollcall(us1890s_votes[[.x]], yea = 1, nay = 0, missing = c(NA, 2, 3)))
us1890s_filtered <- map(1:n_terms, ~filter_votes(rollcall = us1890s_rc[[.x]], lop = 0, minvotes = 20))
us1890s_votes_filtered <- map(1:n_terms, ~as.matrix(us1890s_votes[[.x]])[us1890s_filtered[[.x]]$legis, us1890s_filtered[[.x]]$bills])
us1890s_legis_filtered <- map(1:n_terms, ~us1890s_legis[[.x]][us1890s_filtered[[.x]]$legis, ])
us1890s_bills_filtered <- map(1:n_terms, ~us1890s_bills[[.x]][us1890s_filtered[[.x]]$bills, ])
us1890s_issue_filtered <- map(1:n_terms, ~us1890s_issue[[.x]][us1890s_filtered[[.x]]$bills, ])
```

### Prepare issue label

We can prepare issue labels using `make_issude_code()` function:

```{r issue_label}
us1890s_issue_code <- map(us1890s_issue_filtered, ~make_issue_code(issue_code_vec = .x$issue_label, 
                                                                   level = .x |> 
                                                                         arrange(issue_level) |> 
                                                                         select(issue_label, issue_level) |>
                                                                         distinct() |>
                                                                         pull(issue_label)))
```


## Step 2: Fix the sign-flip

For the identification of the model, we need to fix the sign-flip (please refer to the paper for more details).

We can use the `recode_votes` function to recode votes in the following coding rule: 
- `1` indicates a liberal majority (or conservative minority), and 
- `0` indicates a conservative majority (or liberal minority).

```{r recode_votes}
us1890s_votes_recoded <- map(1:n_terms, ~recode_votes(us1890s_votes_filtered[[.x]],
                                                      party_code = us1890s_legis_filtered[[.x]]$party_name,
                                                      yea = 1,
                                                      nay = 0,
                                                      missing = c(NA, 2, 3),
                                                      liberal_code = "Democrat",
                                                      conservative_code = "Republican",  
                                                      na_threshold = 0.5))
```

Then, we can make dynamic version of `rollcall` object using `make_dynamic_rollcall()` function:

```{r make_dynamic_rollcall}
dynamic_rc <- make_dynamic_rollcall(votes_list = us1890s_votes_recoded,
                      issue_list = us1890s_issue_code,
                      legis_list = us1890s_legis_filtered,
                      colname_legis = "icpsr",
                      colname_party = "group",
                      bills_list = us1890s_bills_filtered,
                      colname_bills = "rollnumber",
                      term_name = c("H52", "H53", "H54"),
                      left_code = 1,
                      right_code = 0,
                      missing_code = NA,
                      notInLegis_code = 9)
```


## Step 3: Starting values

We use Bayesian IRT model to generate the starting values for the `IssueIRT` model.
First, we can use `ideal()` function from the `pscl` package to fit the model.
You may want to increase the number of iterations, thinning, and burn-in based on the convergence diagnostics.

```{r starting_values}
## fit the Bayesian IRT model
set.seed(1)
ideal <- pscl::ideal(dynamic_rc$rollcall, dropList = list(lop = 0, legisMin = 0), # no need to further drop votes and legislators
                     priors = NULL, startvals = "eigen", # default priors and starting values
                     d = 2, maxiter = 500, thin = 1, burnin = 200,
                     impute = FALSE, normalize = FALSE,
                     store.item = TRUE, file = NULL, verbose = FALSE)
```

Recall that Bayesian IRT model has identification problem, and thus we fix $d+1$ number of legislators' ideal points.
We can use `find_pol_rc_horizontal()`, `find_pol_rc_vertical()`, and `find_constraints()` function to select those anchors as follows.

- `find_pol_rc_horizontal()`: Find the most polarized roll call in terms of horizontal division.
- `find_pol_rc_vertical()`: Find the most polarized roll call in terms of vertical division.
- `find_constraints()`: Find the anchors based on the most polarized roll calls.

Alternatively, you may use manually selected legislators using domain knowledge, or pick the legislators based on the visualization of the fitted Bayesian IRT model.

```{r find_anchors}
pol_rc1 <- find_pol_rc_horizontal(dynamic_rc$rollcall, party_code_col = "group", 
                                  liberal_code = "Democrat", conservative_code = "Republican", 
                                  na_threshold = 0.5)
pol_rc2 <- find_pol_rc_vertical(ideal, dynamic_rc$rollcall, pol_rc1, 
                                party_code_col = "group", 
                                na_threshold = 0.5, lop_threshold = 0.1)
const_ls <- find_constraints(ideal, dynamic_rc$rollcall, pol_rc1 = pol_rc1, pol_rc2 = pol_rc2, 
                             party_code_col = "group", left_party_code = "Democrat", 
                             as_list = TRUE)
print(const_ls)
```

Based on these anchors, we post-process the Bayesian IRT model to generate the starting values for the `IssueIRT` model.
We use `postProcess()` function from the `pscl` package.

```{r post_process_birt}
## suppress redundant messages
invisible(capture.output({
  ideal_pp <- pscl::postProcess(ideal, constraints = const_ls)
}))
```

The starting values of ideal points can be visualized as follows. 
Based on this visualization, you may want to adjust the anchors if needed.

```{r visualize_starting_values}
plot_ideal(ideal_point_1d = ideal_pp$xbar[,1], ideal_point_2d = ideal_pp$xbar[,2],
           group = dynamic_rc$rollcall$legis.data$group, p.title = irt.title,
           breaks.group = breaks.group, values.shape = values.shape, values.color = values.color)
```

## Step 4: Make inputs for the model

Now, we are ready to make inputs for the `issueirt_dynamic_stan()` function.
We can use `make_dynamic_stan_input()` function to create the inputs.

```{r make_dynamic_stan_input}
dynamic_stan_input <- make_dynamic_stan_input(dynamic_rollcall = dynamic_rc,
                                      ideal = ideal_pp,
                                      a = 0.01,
                                      b = 0.001,
                                      rho_init = 10,
                                      sd_dynamic = 0.01)
```

## Step 5: Fit the model

Finally, we fit the `IssueIRT` model using the `issueirt_stan` function.
You may want to adjust the number of chains, warmup, and iterations based on the convergence diagnostics and the computational resources.

```{r fit_model, eval=FALSE}
fit_dynamic <- issueirt_dynamic_stan(
  data = dynamic_stan_input$data,    
  init = list(dynamic_stan_input$init, dynamic_stan_input$init), # starting values
  chains = 2,             # number of Markov chains
  warmup = 10,            # number of warmup iterations per chain
  iter = 20,              # total number of iterations per chain
  cores = 2,              # number of cores (could use one per chain)
)
```

```{r call_fit, show=FALSE}
## access the internal dataset
## script is available at data-raw/us_southern_dynamic.R, which is same as the codes above
fit_dynamic <- issueirt:::fit_dynamic
```

## Step 6: Post-process the model parameters

After fitting the model, we can post-process the model to visualize the estimated issue-specific ideal points.
As aforementioned, the issue-specific ideal points are invariant to rotational transformations.
However, the multidimensional ideal points are not invariant to rotational transformations as it is in other IRT models, and thus we need to fix this identification problem by using the anchors, for example.

We can post-process the model using the user-supplied anchors with `get_dynamic_constraints()` function and compute the posterior summaries of the estimated parameters as follows.

```{r post_process}
anchors <- map(names(const_ls), ~str_subset(dynamic_stan_input$misc$legis_term, .x) %>% .[1]) |> unlist()
print(anchors)
const_dynamic <- get_dynamic_constraints(fit_dynamic, dynamic_stan_input, anchors)
posterior_summary_pp <- make_posterior_summary_postprocessed(
  stan_fit = fit_dynamic, 
  constraints = const_dynamic, 
  issue_label = dynamic_stan_input$misc$issue_code_df$codebook$label,
  rc_label = dynamic_stan_input$misc$bills_df$term_rollnumber, 
  legis_label = dynamic_stan_input$misc$legis_term, 
  missing_label = NULL
)
```

The main quantities are as follows:
- `posterior_summary_pp$x_postprocessed`: A tibble of the estimated ideal points of legis-term.
- `posterior_summary_pp$theta_postprocessed`: A tibble of the estimated issue vectors (mean direction, $\theta_z$).
- `posterior_summary_pp$u_postprocessed`: A tibble of the estimated roll call vectors ($u_j$).
Each tibble contains the posterior mean, standard deviation, and quantiles of the estimated parameters.

### Visualization

Now, we visualize the estimated parameters using its posterior mean.

```{r visualize_posterior}
posterior_df_pp <- tibble(legis_term = dynamic_stan_input$misc$legis_term,
                          ideal_point_1d = posterior_summary_pp$x_postprocessed |> filter(dimension == 1) |> pull(mean),
                          ideal_point_2d = posterior_summary_pp$x_postprocessed |> filter(dimension == 2) |> pull(mean),
                          ideal_point_1d_se = posterior_summary_pp$x_postprocessed |> filter(dimension == 1) |> pull(sd),
                          ideal_point_2d_se = posterior_summary_pp$x_postprocessed |> filter(dimension == 2) |> pull(sd)) %>%
  separate(legis_term, into = c("icpsr", "term"), sep = "_") %>%
  mutate(congress = c(52:54)[as.numeric(term)],
         icpsr = as.numeric(icpsr)) %>%
  left_join(dynamic_stan_input$misc$legis_df, by = c("icpsr", "congress"))

plot_ideal(ideal_point_1d = posterior_df_pp$ideal_point_1d, 
           ideal_point_2d = posterior_df_pp$ideal_point_2d,
           group = posterior_df_pp$group, p.title = issueirt.title,
           breaks.group = breaks.group, values.shape = values.shape, values.color = values.color)
```

We can also visualize issue specific axes using `plot_issueaxis()` function.

```{r visualize_issueirt, fig.width=10, fig.height=5}
p_ls <- plot_issueaxis(stan_input = dynamic_stan_input, 
                       posterior_summary = posterior_summary_pp,
                       group = posterior_df_pp$group, p.title = "IssueIRT",
                       breaks.group = breaks.group, values.shape = values.shape, values.color = rep("grey", length(breaks.group)))
library(patchwork)
p_ls[["H52_Monetary"]] + p_ls[["H53_Monetary"]] + p_ls[["H54_Monetary"]] +
  plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position='bottom')
```

Here, each figure visualizes monetary policy specific axis with the following elements:
- Points: multidimensional ideal points
- Black line: issue specific axis
- Grey lines: roll call axes in such issue area ($j$ such that $z_j = z$)
- Dark orange arrow: direction of issue specific axis ($\theta_z$)
- Light orange arrows: direction of roll call axes ($u_j$)

One thing to note is that the direction of some roll call axes in the 52nd and 53rd are in opposite direction to the issue specific axis. 
You may want to revisit step 2 to recode those roll calls, and then proceed with the following steps.


## Step 7: Compute posterior statistics

We can compute the posterior statistics to assess the convergence of the model.
As a warm-up, we start by computing the computation time.

```{r computation}
library(rstan)
elapsed_times <- get_elapsed_time(fit_dynamic)
## convert into minutes
elapsed_times_min <- elapsed_times / 60
## create a data frame for the table
summary_elapsed_times <- data.frame(
  Chain = 1:nrow(elapsed_times_min),
  Warmup = round(elapsed_times_min[, "warmup"], 2),
  Sampling = round(elapsed_times_min[, "sample"], 2)
)
## print table
summary_elapsed_times |> knitr::kable()
```

We can compute $\hat{R}$ statistics and effective sample size (ESS) using functions from `posterior` package, for example.

```{r posterior_statstics_helper}
## helper function to summarize draws
#' function to summarize the convergence statistics
#' @param draws_summary a data frame of draws summary
#' @return a tibble
calculate_convergence_summary <- function(draws_summary) {
  tibble(
    `Min of $\\hat{R}$` = quantile(draws_summary$rhat, 0),
    `25\\% of $\\hat{R}$` = quantile(draws_summary$rhat, 0.25),
    `Median of $\\hat{R}$` = quantile(draws_summary$rhat, 0.5),
    `75\\% of $\\hat{R}$` = quantile(draws_summary$rhat, 0.75),
    `Max of $\\hat{R}$` = quantile(draws_summary$rhat, 1),
    `Mean of $\\hat{R}$` = mean(draws_summary$rhat),
    `Min of ESS` = quantile(draws_summary$ess_bulk, 0),
    `25\\% of ESS` = quantile(draws_summary$ess_bulk, 0.25),
    `Median of ESS` = quantile(draws_summary$ess_bulk, 0.5),
    `75\\% of ESS` = quantile(draws_summary$ess_bulk, 0.75),
    `Max of ESS` = quantile(draws_summary$ess_bulk, 1),
    `Mean of ESS` = mean(draws_summary$ess_bulk)
  )
}
```


```{r posterior_statstics}
## get posterior samples (not just summary as in the previous step)
posterior_samples <- post_process(stan_fit = fit_dynamic, constraints = const_dynamic, legis_label = dynamic_stan_input$misc$legis_term, as_mcmc = TRUE)
posterior_samples_list <- list(
  x1 = posterior_samples$x1,
  x2 = posterior_samples$x2,
  theta = posterior_samples$theta,
  u = posterior_samples$u
)

## convert to draws array format and summarize
summarized_draws <- posterior_samples_list %>%
  map(~ posterior::summarise_draws(posterior::as_draws_array(.)))

## quantiles of statistics for each type of parameters
quantiles <- summarized_draws|>
  map_df(calculate_convergence_summary, .id = "Parameter") |>
  mutate(Parameter = case_when(
    Parameter == "x1" ~ "$x_{i1}$",
    Parameter == "x2" ~ "$x_{i2}$",
    Parameter == "theta" ~ "$\\theta_{z}$",
    Parameter == "u" ~ "$u_{j}$"
  ))

## print results in table
quantiles |> select(Parameter, contains("\\hat{R}")) |> knitr::kable(digits = 2)
quantiles |> select(Parameter, contains("ESS")) |> knitr::kable()
```

We may want to adjust the number of chains, warmup, and iterations based on these convergence diagnostics.


## Step 8: Calculate issue-specific ideal points

Finally, using the fitted parameters from the `IssueIRT` model, we can calculate the issue-specific ideal points with the `get_ideal_points()` function.
We can use `plot_issueirt()` function to visualize the issue-specific ideal points.

```{r calculate_issue_specific, fig.width=10, fig.height=5}
issue_irt <- get_ideal_points(stan_fit = fit_dynamic, 
                              issue_label = dynamic_stan_input$misc$issue_code_df$codebook$label, 
                              legis_label = dynamic_stan_input$misc$legis_term, 
                              legis_group = posterior_df_pp$group)
plot_issueirt(issueirt = issue_irt |> filter(issue_label %in% c("H52_Monetary", "H53_Monetary", "H54_Monetary")), 
              p.title = "Monetary Policy Specific Ideal Points",
              breaks.group = breaks.group, values.shape = values.shape, values.color = values.color) +
  geom_point(alpha = 0.1)
```
